Reduced plasticity and variance in physiological rates for terrestrial ectotherm populations under climate change

### 2024-02-06 ms.qmd
Small update

Climate change is expected to result in warmer and more variable thermal environments globally. Greater thermal variability is expected to result in strong selection pressures leading to genetic adaptation and/or the evolution of adaptive phenotypic plasticity. Such responses depend on genetic and phenotypic variability, yet most work has focused on plastic changes in mean phenotypic responses to climate warming ignore how temperature may change phenotypic variability. Using a large database of physiological rate data from marine, freshwater and terrestrial ectotherms we quantify the capacity for temperature acclimation and the extent to which relative variance (i.e., coefficient of variation) in physiological rates change at high temperatures. Using new effect size estimates and meta-analysis, we show that across habitats relative variance in physiological rates decreased at higher temperatures. Freshwater ectotherms are capable of acclimating and have the smallest reductions in relative variance. In contrast, acclimation responses of marine organisms coincide with a nearly equal reduction in relative variance in physiological rates. Relative variance reductions are particularly pronounced for terrestrial ectotherms, and this coincides with a lack of capacity for acclimation highlighting the vulnerability of terrestrial ectotherms to climate change. Neither life-history stage nor past climate explained effect variability. Our results emphasise how beneficial acclimation responses may be offset by reductions in physiological rate variance that may have important evolutionary and ecological ramifications, challenging the notion that plasticity will be sufficient to counteract climate change effects on populations.

>>>>

Climate change is expected to result in warmer and more variable thermal environments globally. Greater thermal variability is expected to result in strong selection pressures leading to genetic adaptation and/or the evolution of adaptive phenotypic plasticity. Such responses depend on genetic and phenotypic variability, yet most work has focused on plastic changes in mean phenotypic responses to climate warming ignore how temperature may change phenotypic variability. Using a large database of physiological rate data from marine, freshwater and terrestrial ectotherms we quantify the capacity for temperature acclimation and the extent to which relative variance (i.e., coefficient of variation) in physiological rates change at high temperatures. Using new effect size estimates and meta-analysis, we show that across habitats relative variance in physiological rates decreased at higher temperatures. Freshwater ectotherms are capable of acclimating and have the smallest reductions in relative variance. In contrast, acclimation responses of marine organisms coincide with a nearly equal reduction in relative variance in physiological rates. Relative variance reductions are particularly pronounced for terrestrial ectotherms, and this coincides with a lack of capacity for acclimation highlighting the vulnerability of terrestrial ectotherms to climate change. Neither life-history stage nor past climate explained effect variability. Our results emphasise how beneficial acclimation responses may be offset by reductions in physiological rate variance that may have important evolutionary and ecological ramifications challenging the notion that plasticity will be sufficient to counteract climate change effects on populations.
